---
title: "Hands-on_Ex09a"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
execute:
  warning: false
  message: false
---

# Choropleth Mapping with R

## Overview

A choropleth map shades geographic areas according to a data value. It is useful when the variable belongs to an area, such as population by subzone, rate by district or count by planning area.

In this hands-on exercise, Singapore subzone geometry is joined with resident population data. The focus is on importing geospatial and aspatial data, preparing a spatial join key, building choropleth maps and comparing different classification methods.

By the end of this exercise, we should be able to:

- import shapefile data using `sf`;
- import population attribute data using `readr`;
- join attribute values to map polygons;
- build choropleth maps with `ggplot2`;
- compare equal interval, quantile and Jenks classification;
- map selected areas that meet a clear criterion.

## Installing and launching R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  sf,
  classInt,
  scales,
  patchwork,
  knitr
)
```

## Importing Data into R

### The Data

Two data sources are used:

- `MP14_SUBZONE_WEB_PL.shp` contains Singapore Master Plan 2014 subzone boundaries.
- `respopagesextod2011to2020.csv` contains population records by planning area, subzone, age group, sex, type of dwelling and year.

### Importing geospatial data into R

```{r}
shp_path <- "data/geospatial/MP14_SUBZONE_WEB_PL.shp"

if (!file.exists(shp_path)) {
  stop("The subzone shapefile is missing from data/geospatial.")
}

mpsz <- st_read(shp_path, quiet = TRUE) %>%
  st_make_valid()

glimpse(mpsz)
```

```{r}
st_crs(mpsz)
```

### Importing attribute data into R

```{r}
pop_path <- "data/aspatial/respopagesextod2011to2020.csv"

if (!file.exists(pop_path)) {
  stop("The population CSV is missing from data/aspatial.")
}

pop_raw <- read_csv(pop_path, show_col_types = FALSE)

glimpse(pop_raw)
```

### Data Preparation

The map uses total population in 2020. The population file is aggregated by planning area and subzone, then joined to the subzone polygons.

```{r}
pop_2020 <- pop_raw %>%
  filter(Time == 2020) %>%
  group_by(
    PA = str_to_upper(PA),
    SZ = str_to_upper(SZ)
  ) %>%
  summarise(Pop = sum(Pop, na.rm = TRUE), .groups = "drop")

mpsz_clean <- mpsz %>%
  mutate(
    PLN_AREA_N = str_to_upper(PLN_AREA_N),
    SUBZONE_N = str_to_upper(SUBZONE_N)
  )

mpszpop <- mpsz_clean %>%
  left_join(
    pop_2020,
    by = c("PLN_AREA_N" = "PA", "SUBZONE_N" = "SZ")
  ) %>%
  mutate(
    Pop = replace_na(Pop, 0),
    area_sqkm = as.numeric(st_area(.)) / 1000000,
    pop_density = if_else(area_sqkm > 0, Pop / area_sqkm, NA_real_)
  )
```

```{r}
mpszpop %>%
  st_drop_geometry() %>%
  select(PLN_AREA_N, SUBZONE_N, REGION_N, Pop, area_sqkm, pop_density) %>%
  arrange(desc(Pop)) %>%
  head(10) %>%
  knitr::kable(format = "html", digits = 1)
```

## Choropleth Mapping Geospatial Data

### Plotting a choropleth map quickly

::: {.panel-tabset}

## The plot

```{r fig.width=8, fig.height=7}
ggplot(mpszpop) +
  geom_sf(aes(fill = Pop), colour = "white", linewidth = 0.08) +
  scale_fill_gradient(
    low = "#F2F0E6",
    high = "#2A6F6F",
    labels = comma
  ) +
  labs(
    title = "Singapore Resident Population by Subzone, 2020",
    fill = "Population"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(mpszpop) +
  geom_sf(aes(fill = Pop), colour = "white", linewidth = 0.08) +
  scale_fill_gradient(
    low = "#F2F0E6",
    high = "#2A6F6F",
    labels = comma
  ) +
  labs(
    title = "Singapore Resident Population by Subzone, 2020",
    fill = "Population"
  ) +
  theme_void(base_size = 12)
```

:::

This first map uses a continuous colour scale. It is simple, but it can hide important differences when the variable is skewed.

### Data classification methods

Different classification methods can lead to different visual impressions. The helper function below creates map classes using `classInt`.

```{r}
make_class <- function(x, style, n = 5) {
  values <- x[!is.na(x)]
  breaks <- classIntervals(values, n = n, style = style)$brks %>%
    unique()

  cut(
    x,
    breaks = breaks,
    include.lowest = TRUE,
    dig.lab = 10
  )
}

mpszpop <- mpszpop %>%
  mutate(
    equal_class = make_class(Pop, "equal"),
    quantile_class = make_class(Pop, "quantile"),
    jenks_class = make_class(Pop, "jenks")
  )
```

```{r}
plot_choro_class <- function(data, class_col, title) {
  ggplot(data) +
    geom_sf(aes(fill = .data[[class_col]]), colour = "white", linewidth = 0.06) +
    scale_fill_brewer(palette = "YlGnBu", na.value = "grey90") +
    labs(title = title, fill = NULL) +
    theme_void(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      legend.position = "bottom",
      legend.text = element_text(size = 6)
    )
}
```

```{r fig.width=11, fig.height=8}
plot_choro_class(mpszpop, "equal_class", "Equal Interval") +
  plot_choro_class(mpszpop, "quantile_class", "Quantile") +
  plot_choro_class(mpszpop, "jenks_class", "Jenks Natural Breaks")
```

The same population values are used in all three maps. The change in appearance comes from the classification method, not from the underlying data.

### Colour Scheme

Sequential palettes work well for ordered numerical values. In the map below, a single-hue scale is used for population density so dense areas stand out.

```{r fig.width=8, fig.height=7}
ggplot(mpszpop) +
  geom_sf(aes(fill = pop_density), colour = "white", linewidth = 0.08) +
  scale_fill_viridis_c(
    option = "mako",
    labels = label_number(scale_cut = cut_short_scale()),
    na.value = "grey90"
  ) +
  labs(
    title = "Population Density by Subzone, 2020",
    subtitle = "Density adjusts population count by polygon area",
    fill = "People per sq km"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Mapping spatial objects meeting a selection criterion

The map below highlights subzones in the top 10 percent of population density. This is a small extension from a general choropleth map to a targeted selection map.

```{r}
density_cutoff <- quantile(mpszpop$pop_density, 0.90, na.rm = TRUE)

dense_subzones <- mpszpop %>%
  mutate(
    density_group = if_else(pop_density >= density_cutoff, "Top 10%", "Other subzones")
  )
```

```{r fig.width=8, fig.height=7}
ggplot(dense_subzones) +
  geom_sf(aes(fill = density_group), colour = "white", linewidth = 0.08) +
  scale_fill_manual(
    values = c("Top 10%" = "#D95F02", "Other subzones" = "grey88")
  ) +
  labs(
    title = "Subzones with High Population Density",
    subtitle = "Highlighted areas are in the top 10 percent of density",
    fill = NULL
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## My Extension: Planning area summary

The map tells us where population is concentrated, but a table makes it easier to identify the largest planning areas.

```{r}
mpszpop %>%
  st_drop_geometry() %>%
  group_by(PLN_AREA_N, REGION_N) %>%
  summarise(
    population = sum(Pop, na.rm = TRUE),
    area_sqkm = sum(area_sqkm, na.rm = TRUE),
    density = population / area_sqkm,
    .groups = "drop"
  ) %>%
  arrange(desc(population)) %>%
  head(12) %>%
  mutate(
    population = comma(population),
    area_sqkm = round(area_sqkm, 2),
    density = comma(round(density))
  ) %>%
  knitr::kable(format = "html")
```

## Key Takeaways

- Choropleth mapping requires a reliable join between attribute data and polygon geometry.
- Classification method affects interpretation, so the choice should match the analytical purpose.
- Population density can tell a different story from total population because polygon areas differ.
- Selection maps are useful when the goal is to highlight areas that meet a condition.

## References

- R4VA chapter: <https://r4va.netlify.app/chap21>

