---
title: "Hands-on_Ex09b"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
execute:
  warning: false
  message: false
---

# Visualising Geospatial Point Data

## Overview

Point maps are used when each observation has a coordinate pair. Unlike choropleth maps, the unit of analysis is a location rather than an area. Symbol size, colour and shape can then be used to encode additional attributes.

In this hands-on exercise, Singapore Pools outlet data is converted from an aspatial table into an `sf` point layer. The points are mapped over Singapore subzones and then redesigned as proportional symbol maps.

By the end of this exercise, we should be able to:

- import point data with projected coordinates;
- convert a data frame into an `sf` point object;
- draw point symbol maps;
- draw proportional symbol maps;
- compare point patterns by outlet type;
- use a base polygon layer to provide geographic context.

## Installing and launching R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  sf,
  scales,
  knitr
)
```

## Geospatial Data Wrangling

### The data

The point data file is `SGPools_svy21.csv`. It includes outlet names, addresses, coordinates, outlet type and the `Gp1Gp2 Winnings` field used for proportional symbols.

A Singapore subzone shapefile is also included in this exercise folder so the points can be mapped on a recognisable base map.

```{r}
pools_path <- "data/aspatial/SGPools_svy21.csv"
sg_path <- "data/geospatial/MP14_SUBZONE_WEB_PL.shp"

if (!file.exists(pools_path) || !file.exists(sg_path)) {
  stop("The Singapore Pools CSV or Singapore subzone shapefile is missing.")
}

sg_map <- st_read(sg_path, quiet = TRUE) %>%
  st_make_valid()

sgpools <- read_csv(pools_path, show_col_types = FALSE)

glimpse(sgpools)
```

### Creating a sf data frame from an aspatial data frame

The coordinates are stored in Singapore SVY21 projected coordinates. The point layer is created using the same CRS as the subzone map to avoid misalignment.

```{r}
sgpools_sf <- st_as_sf(
  sgpools,
  coords = c("XCOORD", "YCOORD"),
  crs = st_crs(sg_map),
  remove = FALSE
) %>%
  mutate(
    winnings = `Gp1Gp2 Winnings`,
    outlet_type = `OUTLET TYPE`
  )
```

```{r}
sgpools_sf %>%
  st_drop_geometry() %>%
  select(NAME, outlet_type, winnings, XCOORD, YCOORD) %>%
  head(10) %>%
  knitr::kable(format = "html")
```

## Drawing Proportional Symbol Map

### It all started with an interactive point symbol map

The first map uses equal-sized points. It is useful for checking whether the points are placed correctly.

::: {.panel-tabset}

## The plot

```{r fig.width=8, fig.height=7}
ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(data = sgpools_sf, aes(colour = outlet_type), size = 2.2, alpha = 0.85) +
  coord_sf() +
  labs(
    title = "Singapore Pools Outlet Locations",
    colour = "Outlet type"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## The code chunk

```{r eval=FALSE}
ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(data = sgpools_sf, aes(colour = outlet_type), size = 2.2, alpha = 0.85) +
  coord_sf() +
  labs(
    title = "Singapore Pools Outlet Locations",
    colour = "Outlet type"
  ) +
  theme_void(base_size = 12)
```

:::

### Lets make it proportional

The next map uses point size to represent `Gp1Gp2 Winnings`. This is more informative because it shows both location and magnitude.

```{r fig.width=8, fig.height=7}
ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(
    data = sgpools_sf,
    aes(size = winnings),
    shape = 21,
    fill = "#2A6F6F",
    colour = "white",
    alpha = 0.82,
    stroke = 0.25
  ) +
  scale_size_area(max_size = 12, labels = comma) +
  labs(
    title = "Proportional Symbol Map of Singapore Pools Outlets",
    subtitle = "Symbol area represents Group 1 and Group 2 winnings",
    size = "Winnings"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

### Lets give it a different colour

Colour can be used together with size to represent outlet type and winnings at the same time.

```{r fig.width=8, fig.height=7}
ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(
    data = sgpools_sf,
    aes(size = winnings, fill = outlet_type),
    shape = 21,
    colour = "white",
    alpha = 0.86,
    stroke = 0.25
  ) +
  scale_size_area(max_size = 12, labels = comma) +
  scale_fill_manual(values = c("#2A6F6F", "#D95F02", "#F2A65A", "#7570B3")) +
  labs(
    title = "Outlet Type and Winnings",
    size = "Winnings",
    fill = "Outlet type"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

### I have a twin brother

Faceting gives each outlet type its own panel. This reduces overlap and makes the spatial pattern of each group easier to inspect.

```{r fig.width=10, fig.height=7}
ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(
    data = sgpools_sf,
    aes(size = winnings),
    shape = 21,
    fill = "#D95F02",
    colour = "white",
    alpha = 0.85,
    stroke = 0.2
  ) +
  facet_wrap(~outlet_type) +
  scale_size_area(max_size = 9, labels = comma) +
  labs(
    title = "Proportional Symbols by Outlet Type",
    size = "Winnings"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## My Extension: Highest winning outlets

The table below identifies the outlets with the largest recorded winnings. This provides a readable supplement to the proportional symbol map, where large circles can overlap.

```{r}
sgpools_sf %>%
  st_drop_geometry() %>%
  arrange(desc(winnings)) %>%
  select(NAME, ADDRESS, outlet_type, winnings) %>%
  head(10) %>%
  mutate(winnings = comma(winnings)) %>%
  knitr::kable(format = "html")
```

```{r fig.width=8, fig.height=7}
top_outlets <- sgpools_sf %>%
  slice_max(winnings, n = 10, with_ties = FALSE)

ggplot() +
  geom_sf(data = sg_map, fill = "grey94", colour = "white", linewidth = 0.08) +
  geom_sf(data = sgpools_sf, colour = "grey55", size = 1.3, alpha = 0.45) +
  geom_sf(
    data = top_outlets,
    aes(size = winnings),
    shape = 21,
    fill = "#D95F02",
    colour = "white",
    alpha = 0.9
  ) +
  scale_size_area(max_size = 13, labels = comma) +
  labs(
    title = "Top Winning Outlets",
    subtitle = "Highlighted points are the ten outlets with the largest winnings",
    size = "Winnings"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Key Takeaways

- Point maps require correctly assigned coordinate reference systems.
- Equal-sized point maps are good for checking location, while proportional symbols add magnitude.
- Symbol size and colour should not compete; each should encode a clear variable.
- A summary table is helpful when overlapping circles make exact comparison difficult.

## References

- R4VA chapter: <https://r4va.netlify.app/chap22>
