---
title: "Hands-on_Ex07b"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
execute:
  warning: false
  message: false
---

# Visualising and Analysing Text Data with R

## Overview

Text data is unstructured, so it needs to be reshaped before it can be counted, compared and visualised. The tidy text approach converts documents into one-token-per-row data, making it possible to use familiar `dplyr` and `ggplot2` workflows.

In this hands-on exercise, the 20 Newsgroups sample is used to practise text import, cleaning, tokenisation, word frequency analysis, TF-IDF, word correlation and bigram network visualisation.

By the end of this exercise, we should be able to:

- import multiple text files from multiple folders;
- clean email-style message text;
- tokenise text into individual words and bigrams;
- remove stop words and noisy tokens;
- compute term frequency and TF-IDF;
- visualise word co-occurrence and bigram relationships as networks.

## Installing and launching R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  tidytext,
  widyr,
  tidygraph,
  ggraph,
  igraph,
  DT,
  scales,
  knitr
)
```

## Importing multiple text files from multiple folders

### Creating a folder list

The data folder contains one subfolder per newsgroup. Each file inside a subfolder is one message.

```{r}
news_root <- "data/20news"

if (!dir.exists(news_root)) {
  stop("The folder data/20news is missing. Please unzip 20news.zip into the data folder.")
}

news_folders <- list.dirs(news_root, recursive = FALSE, full.names = TRUE)

tibble(
  folder = basename(news_folders),
  messages = map_int(news_folders, ~ length(list.files(.x, full.names = TRUE)))
) %>%
  knitr::kable(format = "html")
```

### Define a function to read all files from a folder

```{r}
read_newsgroup <- function(folder) {
  files <- list.files(folder, full.names = TRUE)

  tibble(
    newsgroup = basename(folder),
    file = basename(files),
    text = map_chr(files, readr::read_file)
  )
}
```

### Reading in all messages

```{r}
news_raw <- map_dfr(news_folders, read_newsgroup) %>%
  mutate(doc_id = row_number())

glimpse(news_raw)
```

```{r}
news_raw %>%
  count(newsgroup, sort = TRUE) %>%
  knitr::kable(format = "html")
```

## Initial EDA

Before tokenising, it is useful to inspect message length. Very short messages may contain only quoted text or metadata, while very long messages may dominate word counts.

```{r}
news_eda <- news_raw %>%
  mutate(
    characters = str_length(text),
    lines = str_count(text, "\n") + 1
  )
```

```{r fig.width=8, fig.height=5}
news_eda %>%
  ggplot(aes(characters)) +
  geom_histogram(bins = 30, fill = "#2A6F6F", colour = "white") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Distribution of Message Length",
    x = "Number of characters",
    y = "Number of messages"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Introducing tidytext

### Removing header and quoted lines

Newsgroup files often include headers, signatures and quoted replies. The cleaning function below removes the first header block, removes quoted lines beginning with `>`, and squeezes repeated whitespace.

```{r}
clean_message <- function(x) {
  x %>%
    str_replace_all("\r\n", "\n") %>%
    str_replace(regex("^.*?\n\n", dotall = TRUE), "") %>%
    str_replace_all(regex("^>.*$", multiline = TRUE), " ") %>%
    str_replace_all(regex("^\\|.*$", multiline = TRUE), " ") %>%
    str_replace_all("\\b[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[A-Za-z]{2,}\\b", " ") %>%
    str_replace_all("[^A-Za-z\\s]", " ") %>%
    str_squish()
}

news_clean <- news_raw %>%
  mutate(text_clean = map_chr(text, clean_message))
```

### Text data processing

The `unnest_tokens()` function converts each message into one word per row. Common stop words are removed so the remaining terms are more meaningful.

```{r}
data("stop_words")

news_words <- news_clean %>%
  select(doc_id, newsgroup, text_clean) %>%
  unnest_tokens(word, text_clean) %>%
  anti_join(stop_words, by = "word") %>%
  filter(
    str_detect(word, "^[a-z]+$"),
    str_length(word) > 2
  )

glimpse(news_words)
```

### Visualising words in newsgroups

```{r}
word_counts <- news_words %>%
  count(newsgroup, word, sort = TRUE)

focus_groups <- news_clean %>%
  count(newsgroup, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(newsgroup)

top_words <- word_counts %>%
  filter(newsgroup %in% focus_groups) %>%
  group_by(newsgroup) %>%
  slice_max(n, n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, n, newsgroup))
```

```{r fig.width=10, fig.height=7}
ggplot(top_words, aes(n, word, fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~newsgroup, scales = "free_y") +
  labs(
    title = "Most Frequent Words by Newsgroup",
    x = "Word count",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

## Basic concept of TF-IDF

Frequent words are not always distinctive. TF-IDF downweights terms that appear in many groups and highlights words that are unusually important to a particular group.

### Computing TF-IDF within newsgroups

```{r}
tf_idf_words <- word_counts %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

```{r}
DT::datatable(
  tf_idf_words %>%
    select(newsgroup, word, n, tf, idf, tf_idf) %>%
    head(60),
  rownames = FALSE,
  options = list(pageLength = 10)
)
```

### Visualising TF-IDF within newsgroups

```{r}
top_tfidf <- tf_idf_words %>%
  filter(newsgroup %in% focus_groups) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, n = 8, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, tf_idf, newsgroup))
```

```{r fig.width=10, fig.height=7}
ggplot(top_tfidf, aes(tf_idf, word, fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  scale_y_reordered() +
  facet_wrap(~newsgroup, scales = "free_y") +
  labs(
    title = "Distinctive Terms by Newsgroup",
    subtitle = "TF-IDF highlights words that are frequent in one group but less common elsewhere",
    x = "TF-IDF",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

## Counting and correlating pairs of words

The `widyr` package can calculate how often two words appear in the same document. This is useful for turning text into a network.

```{r}
word_correlation <- news_words %>%
  add_count(word) %>%
  filter(n >= 5) %>%
  pairwise_cor(word, doc_id, sort = TRUE, upper = FALSE) %>%
  filter(correlation >= 0.25) %>%
  slice_head(n = 80)

word_correlation %>%
  head(10) %>%
  knitr::kable(format = "html", digits = 3)
```

```{r fig.width=9, fig.height=7}
if (nrow(word_correlation) > 0) {
  word_graph <- word_correlation %>%
    transmute(from = item1, to = item2, correlation) %>%
    as_tbl_graph(directed = FALSE)

  set.seed(608)

  ggraph(word_graph, layout = "fr") +
    geom_edge_link(aes(width = correlation), alpha = 0.35, colour = "grey55") +
    geom_node_point(size = 3, colour = "#2A6F6F") +
    geom_node_text(aes(label = name), size = 3, vjust = 1.6, check_overlap = TRUE) +
    scale_edge_width(range = c(0.3, 2), guide = "none") +
    labs(
      title = "Word Correlation Network",
      subtitle = "Words are linked when they tend to appear in the same messages"
    ) +
    theme_graph(base_family = "sans") +
    theme(plot.title = element_text(face = "bold"))
}
```

## Bigram analysis

Bigrams are adjacent two-word phrases. They preserve a small amount of word order, which can make the text interpretation more meaningful than individual tokens alone.

```{r}
news_bigrams <- news_clean %>%
  select(doc_id, newsgroup, text_clean) %>%
  unnest_tokens(bigram, text_clean, token = "ngrams", n = 2) %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(
    !word1 %in% stop_words$word,
    !word2 %in% stop_words$word,
    str_detect(word1, "^[a-z]+$"),
    str_detect(word2, "^[a-z]+$"),
    str_length(word1) > 2,
    str_length(word2) > 2
  )

bigram_counts <- news_bigrams %>%
  count(word1, word2, sort = TRUE)

bigram_counts %>%
  head(12) %>%
  knitr::kable(format = "html")
```

### Visualising a network of bigrams with ggraph

```{r}
bigram_graph <- bigram_counts %>%
  slice_max(n, n = 80, with_ties = FALSE) %>%
  transmute(from = word1, to = word2, weight = n) %>%
  as_tbl_graph(directed = TRUE)
```

```{r fig.width=10, fig.height=7}
set.seed(608)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(
    aes(width = weight),
    alpha = 0.35,
    colour = "grey55",
    arrow = arrow(length = unit(2, "mm"), type = "closed"),
    end_cap = circle(2, "mm")
  ) +
  geom_node_point(colour = "#D95F02", size = 3) +
  geom_node_text(aes(label = name), size = 3, vjust = 1.6, check_overlap = TRUE) +
  scale_edge_width(range = c(0.2, 2), guide = "none") +
  labs(
    title = "Bigram Network",
    subtitle = "Directional links show the first and second word in common two-word phrases"
  ) +
  theme_graph(base_family = "sans") +
  theme(plot.title = element_text(face = "bold"))
```

## My Extension: Message length by newsgroup

This extension checks whether some groups tend to write longer messages. It is a useful caution because longer documents naturally generate more words and more possible co-occurrences.

```{r fig.width=10, fig.height=6}
news_eda %>%
  mutate(newsgroup = fct_reorder(newsgroup, characters, median)) %>%
  ggplot(aes(characters, newsgroup)) +
  geom_boxplot(fill = "#E6B655", colour = "grey35", outlier.alpha = 0.35) +
  scale_x_log10(labels = comma) +
  labs(
    title = "Message Length Varies by Newsgroup",
    x = "Characters per message, log scale",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Key Takeaways

- Tidy text analysis begins by reshaping documents into token-level rows.
- Removing stop words and obvious message metadata improves the quality of word counts.
- TF-IDF is useful when the question is about distinctive vocabulary, not only frequent vocabulary.
- Word and bigram networks reveal associations that are hard to see in a simple frequency table.

## References

- R4VA chapter: <https://r4va.netlify.app/chap29>

