---
title: "Hands-on Exercise 4: Programming Animated Statistical Graphics with R"
author: "Cheng Yuanyuan"
date: "2026-05-01"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

Static charts are good at showing one frozen moment. Animated statistical graphics are useful when the data has a time dimension and the story depends on movement, change, or progression.

In this exercise, I use R to create animated visualisations. The focus is not to make the chart move for entertainment. Animation should help the reader understand how patterns change over time. If the movement does not add meaning, then it is just a chart doing cardio for no reason.

The dataset used in this exercise contains country-level population, GDP per capita, life expectancy and regional information over time. This makes it suitable for animated bubble charts and animated trend charts.

The guiding question for this exercise is:

> Can animation reveal a pattern that would be harder to see in a static chart?

# Getting Started

Before creating the animated charts, I load the packages needed for data wrangling, plotting and animation. In this version, I use `plotly` for animation because it works well in HTML websites and is more stable for Vercel deployment.

```{r}
pacman::p_load(
  tidyverse,
  gapminder,
  plotly,
  scales
)
```

# Importing Data

For this exercise, I use the `gapminder` dataset. It contains country-level data across different years, including life expectancy, GDP per capita, population size and continent.

This structure is suitable for animated statistical graphics because each country can be tracked across time.

```{r}
data("gapminder")

gapminder_data <- gapminder
```

```{r}
glimpse(gapminder_data)
```

# Data Preparation

Before drawing the charts, I prepare a cleaner version of the dataset. Population is converted into millions so that the values are easier to read.

I also create tooltip text for the animated charts. This allows the reader to hover over each point or bar and inspect the country-level details without overcrowding the visualisation.

```{r}
gapminder_clean <- gapminder_data %>%
  mutate(
    population_million = pop / 1000000,
    tooltip = paste0(
      "Country: ", country,
      "<br>Continent: ", continent,
      "<br>Year: ", year,
      "<br>Life expectancy: ", round(lifeExp, 1),
      "<br>GDP per capita: ", dollar(round(gdpPercap, 0)),
      "<br>Population: ", round(population_million, 1), " million"
    )
  )
```

```{r}
gapminder_clean %>%
  select(country, continent, year, lifeExp, gdpPercap, population_million) %>%
  head()
```

# Static Bubble Chart

Before adding animation, I first create a static bubble chart for one selected year. This acts as a visual checkpoint. If the static chart is unclear, animating it will not magically make it better.

In this chart, GDP per capita is shown on the x-axis, life expectancy is shown on the y-axis, bubble size represents population, and colour represents continent.

```{r}
gapminder_2007 <- gapminder_clean %>%
  filter(year == 2007)

ggplot(
  gapminder_2007,
  aes(
    x = gdpPercap,
    y = lifeExp,
    size = population_million,
    colour = continent
  )
) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = dollar) +
  scale_size_continuous(
    range = c(2, 18),
    name = "Population\n(million)"
  ) +
  labs(
    title = "GDP per Capita and Life Expectancy in 2007",
    subtitle = "Bubble size represents population; colour represents continent",
    x = "GDP per Capita, log scale",
    y = "Life Expectancy",
    colour = "Continent"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```

# Animated Bubble Chart

The static bubble chart shows one year, but it cannot show how countries move over time. To reveal the time-based pattern, I now animate the same chart across years.

Each bubble represents one country. The x-axis shows GDP per capita, the y-axis shows life expectancy, bubble size represents population, and colour represents continent. The play button allows the reader to watch how countries move across time.

```{r}
plot_ly(
  data = gapminder_clean,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~population_million,
  color = ~continent,
  frame = ~year,
  ids = ~country,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(
    opacity = 0.7,
    line = list(
      width = 1,
      color = "#1F1F1F"
    )
  )
) %>%
  layout(
    title = "GDP per Capita and Life Expectancy Over Time",
    xaxis = list(
      title = "GDP per Capita, log scale",
      type = "log"
    ),
    yaxis = list(title = "Life Expectancy"),
    legend = list(title = list(text = "Continent"))
  )
```

# Animated Line Chart

The bubble chart is useful for comparing countries, but it can feel visually busy because many bubbles move at the same time. To make the time trend clearer, I create an animated line chart showing average life expectancy by continent.

This chart focuses on one question: how did life expectancy change across continents over time?

```{r}
continent_life <- gapminder_clean %>%
  group_by(continent, year) %>%
  summarise(
    avg_lifeExp = mean(lifeExp),
    .groups = "drop"
  )

plot_ly(
  data = continent_life,
  x = ~year,
  y = ~avg_lifeExp,
  color = ~continent,
  frame = ~year,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste0(
    "Continent: ", continent,
    "<br>Year: ", year,
    "<br>Average life expectancy: ", round(avg_lifeExp, 1)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Average Life Expectancy by Continent Over Time",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average Life Expectancy"),
    legend = list(title = list(text = "Continent"))
  )
```

# Animated Bar Chart

The previous animated charts show movement and trends, but they do not directly show ranking. For ranking changes, an animated bar chart is easier to read.

In this chart, I focus on the ten most populated countries in each year. The bars show population size in millions, and the animation reveals how the population ranking changes over time.

```{r}
top_population <- gapminder_clean %>%
  group_by(year) %>%
  slice_max(
    order_by = population_million,
    n = 10,
    with_ties = FALSE
  ) %>%
  ungroup()

plot_ly(
  data = top_population,
  x = ~population_million,
  y = ~country,
  color = ~continent,
  frame = ~year,
  type = "bar",
  orientation = "h",
  text = ~paste0(
    "Country: ", country,
    "<br>Continent: ", continent,
    "<br>Year: ", year,
    "<br>Population: ", round(population_million, 1), " million"
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "Top 10 Most Populated Countries Over Time",
    xaxis = list(title = "Population (million)"),
    yaxis = list(title = ""),
    legend = list(title = list(text = "Continent"))
  )
```

# Key Observations

The animated bubble chart shows a broad development pattern: many countries move toward higher GDP per capita and higher life expectancy over time. This movement is easier to understand through animation because the reader can follow the direction of change rather than comparing many separate static charts.

The animated line chart gives a cleaner view of life expectancy trends by continent. It shows that life expectancy generally increases over time, although the level and speed of improvement differ across continents.

The animated bar chart is useful for showing population ranking. It gives a direct view of which countries have the largest populations in each year. Compared with the bubble chart, it answers a narrower question, but it answers that question more clearly.

Overall, the three animated charts serve different purposes. The bubble chart is good for multidimensional movement, the line chart is good for long-term trend comparison, and the bar chart is good for ranking changes.

# Conclusion

This exercise shows that animated statistical graphics are useful when the dataset contains a meaningful time dimension. Animation helps the reader follow change, movement and progression instead of comparing many separate static charts.

However, animation should not be used just because it looks modern. A moving chart can still be a bad chart if the movement does not support the message. In this exercise, the animation works because the data naturally changes over time: countries develop, populations grow, and life expectancy shifts across years.

The main lesson is that animation should clarify change. If the movement helps the reader understand a pattern more easily, it adds value. If it only makes the chart look busy, then it is just visual noise with better choreography.