---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics - ggplot2 methods"
author: "Cheng Yuanyuan"
---

# **1.1 Learning Outcome**

In this chapter, you will learn the fundamental principles and key components of **ggplot2**. You will also gain practical experience using these components to create statistical graphics based on the **Layered Grammar of Graphics**. By the end of this chapter, you will be able to apply the essential graphical elements in ggplot2 to produce statistical visualisations that are both elegant and functional.

# 1.2 Getting started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If the are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("Exam_data.csv")

```

## plotting a simple bar chart

```{r}
#｜echo: false
ggplot(data = exam_data,
       aes(x = RACE)) + 
  geom_bar()
```
