---
title: "Hands-on_Ex05a"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
---

# Visualising Distribution

## Learning Outcome

In this exercise, we learn how to visualise data distribution using two methods: ridgeline plot and raincloud plot.

A ridgeline plot is useful when we want to compare the distribution of one numeric variable across many groups. A raincloud plot is useful because it combines density shape, boxplot summary, and individual data points in one chart.

By the end of this exercise, we should be able to:

- create ridgeline plots using `ggridges`;
- create raincloud plots using `ggdist`;
- compare distribution shapes across different groups;
- explain what the chart tells us.

## Getting Started

### Installing and loading the packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

### Importing the data

```{r}
exam <- read_csv("../data/Exam_data.csv")
```

```{r}
glimpse(exam)
```

```{r}
summary(exam$ENGLISH)
```

## Visualising Distribution with Ridgeline Plot

A ridgeline plot is useful when the same numeric variable needs to be compared across multiple groups. In this example, the English score distribution is compared across different classes.

The advantage of this method is that it keeps all classes on one shared x-axis, making it easier to compare the shape, centre and spread of each class distribution.

### Plotting ridgeline graph: ggridges method

The basic ridgeline plot can be created using `geom_density_ridges()`. Here, `ENGLISH` is placed on the x-axis and `CLASS` is placed on the y-axis.

The fill colour is adjusted to a muted teal tone to make the plot look cleaner and more consistent with the overall website style.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 2.8,
    rel_min_height = 0.01,
    bandwidth = 3.2,
    fill = lighten("#2A6F6F", .35),
    color = "white",
    linewidth = 0.35
  ) +
  scale_x_continuous(
    name = "English scores",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.4))
  ) +
  labs(
    title = "Distribution of English Scores by Class",
    subtitle = "Ridgeline plot allows multiple class distributions to be compared together"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 2.8,
    rel_min_height = 0.01,
    bandwidth = 3.2,
    fill = lighten("#2A6F6F", .35),
    color = "white",
    linewidth = 0.35
  ) +
  scale_x_continuous(
    name = "English scores",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.4))
  ) +
  labs(
    title = "Distribution of English Scores by Class",
    subtitle = "Ridgeline plot allows multiple class distributions to be compared together"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

:::

From this plot, we can compare whether some classes have more concentrated English scores while others show wider score variation. A taller or sharper ridge usually suggests that many students are clustered around a similar score range.

### Varying fill colors along the x axis

Instead of filling each ridge with one fixed colour, we can allow the colour to change along the x-axis. This is done using `geom_density_ridges_gradient()`.

This version makes the score direction more visible because lower and higher English scores are represented by different colour intensity.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_c(
    name = "English score",
    option = "C"
  ) +
  scale_x_continuous(
    name = "English scores",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.4))
  ) +
  labs(
    title = "Gradient Ridgeline Plot of English Scores",
    subtitle = "Colour changes along the score axis to highlight lower and higher score ranges"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = after_stat(x))) +
  geom_density_ridges_gradient(
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_c(
    name = "English score",
    option = "C"
  ) +
  scale_x_continuous(
    name = "English scores",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.4))
  ) +
  labs(
    title = "Gradient Ridgeline Plot of English Scores",
    subtitle = "Colour changes along the score axis to highlight lower and higher score ranges"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

:::

This plot is useful when the audience needs to quickly identify where low, middle and high scores appear in each class distribution.

### Mapping the probabilities directly onto colour

Another way to colour a ridgeline plot is to map the empirical cumulative distribution function, also known as ECDF, onto the fill colour. This helps highlight the central region and tail areas of the distribution.

In the code below, `calc_ecdf = TRUE` is important because it tells `stat_density_ridges()` to calculate ECDF-related values.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5 - after_stat(ecdf)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_c(
    name = "Tail probability",
    option = "D",
    direction = -1
  ) +
  labs(
    title = "Ridgeline Plot with Probability-based Colouring",
    subtitle = "The colour scale helps separate central values from distribution tails"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5 - after_stat(ecdf)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_c(
    name = "Tail probability",
    option = "D",
    direction = -1
  ) +
  labs(
    title = "Ridgeline Plot with Probability-based Colouring",
    subtitle = "The colour scale helps separate central values from distribution tails"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

:::

::: {.callout-important}
The argument `calc_ecdf = TRUE` should not be removed. Without it, the ECDF-based colour mapping will not work properly.
:::

### Ridgeline plots with quantile lines

Quantile lines can be added to show how the distribution is divided into different sections. In the example below, the distribution is divided into four quartiles.

This makes the ridgeline plot more informative because it shows not only the density shape, but also the internal score segmentation.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE,
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_d(
    name = "Quartiles",
    option = "C"
  ) +
  labs(
    title = "Ridgeline Plot with Quartile Lines",
    subtitle = "Quartile lines divide each class distribution into four score sections"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE,
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_viridis_d(
    name = "Quartiles",
    option = "C"
  ) +
  labs(
    title = "Ridgeline Plot with Quartile Lines",
    subtitle = "Quartile lines divide each class distribution into four score sections"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

:::

We can also define the quantiles manually. The next chart separates the lower tail, middle range and upper tail using 2.5% and 97.5% cut points.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975),
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#D95F02A0", "#2A6F6FA0", "#7570B3A0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  labs(
    title = "Ridgeline Plot with Tail Probability Cut Points",
    subtitle = "Lower and upper tails are separated from the main score range"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(after_stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975),
    scale = 2.8,
    rel_min_height = 0.01,
    color = "white",
    linewidth = 0.3
  ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#D95F02A0", "#2A6F6FA0", "#7570B3A0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  labs(
    title = "Ridgeline Plot with Tail Probability Cut Points",
    subtitle = "Lower and upper tails are separated from the main score range"
  ) +
  theme_ridges() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9)
  )
```

:::

This version is useful when the analysis is more interested in extreme score ranges, such as unusually low or unusually high English scores.

## Visualising Distribution with Raincloud Plot

A raincloud plot combines three layers:

- a half-density plot showing the distribution shape;
- a boxplot showing median and interquartile range;
- dots showing individual observations.

Compared with a normal boxplot, a raincloud plot gives a fuller view because it shows both summary statistics and the actual data pattern.

In this section, the English score distribution is compared across different race groups.

### Plotting a Half Eye graph

The first layer is created using `stat_halfeye()` from the `ggdist` package. It produces a half-density shape.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  labs(
    title = "Half-eye Plot of English Scores by Race",
    subtitle = "The half-density layer shows the distribution shape for each group",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  labs(
    title = "Half-eye Plot of English Scores by Race",
    subtitle = "The half-density layer shows the distribution shape for each group",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

:::

::: {.callout-tip}
Setting `.width = 0` and `point_colour = NA` removes the interval and point marker, leaving a cleaner half-density shape.
:::

### Adding the boxplot with `geom_boxplot()`

Next, a narrow boxplot is added on top of the half-eye layer. The boxplot provides a compact summary of the median, lower quartile and upper quartile.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  labs(
    title = "Half-eye Plot with Boxplot Layer",
    subtitle = "The boxplot adds a statistical summary to the distribution shape",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  labs(
    title = "Half-eye Plot with Boxplot Layer",
    subtitle = "The boxplot adds a statistical summary to the distribution shape",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

:::

### Adding the Dot Plots with `stat_dots()`

The third layer adds dots to represent individual observations. This makes the plot less abstract because the viewer can see the actual data points behind the distribution.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  stat_dots(
    side = "left",
    justification = 1.15,
    binwidth = .5,
    dotsize = 1.6,
    color = "#1F1F1F",
    fill = "#F2A65A"
  ) +
  labs(
    title = "Raincloud Plot with Dot Layer",
    subtitle = "Dots show the individual observations within each race group",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  stat_dots(
    side = "left",
    justification = 1.15,
    binwidth = .5,
    dotsize = 1.6,
    color = "#1F1F1F",
    fill = "#F2A65A"
  ) +
  labs(
    title = "Raincloud Plot with Dot Layer",
    subtitle = "Dots show the individual observations within each race group",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal()
```

:::

This layer is useful because it prevents the chart from hiding the sample size and the actual spread of observations.

### Finishing touch

Finally, `coord_flip()` is used to turn the chart horizontally. This makes the raincloud shape easier to read and gives more space for group labels.

A clean minimal theme is used here instead of a heavy publication theme, so the chart looks more modern and consistent with the customised colourway.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  stat_dots(
    side = "left",
    justification = 1.15,
    binwidth = .5,
    dotsize = 1.45,
    color = "#1F1F1F",
    fill = "#F2A65A"
  ) +
  coord_flip() +
  labs(
    title = "Raincloud Plot of English Scores by Race",
    subtitle = "Density, boxplot and individual dots are combined in one chart",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(
    adjust = 0.5,
    justification = -0.25,
    .width = 0,
    point_colour = NA,
    fill = lighten("#2A6F6F", .35),
    color = "#1F1F1F"
  ) +
  geom_boxplot(
    width = .18,
    outlier.shape = NA,
    alpha = 0.55,
    color = "#1F1F1F"
  ) +
  stat_dots(
    side = "left",
    justification = 1.15,
    binwidth = .5,
    dotsize = 1.45,
    color = "#1F1F1F",
    fill = "#F2A65A"
  ) +
  coord_flip() +
  labs(
    title = "Raincloud Plot of English Scores by Race",
    subtitle = "Density, boxplot and individual dots are combined in one chart",
    x = "Race",
    y = "English scores"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

### My Extension: Comparing Central Tendency by Race

To add a small extension beyond the original tutorial, we can calculate the median English score for each race group. This provides a simple numerical check to support the visual comparison made in the raincloud plot.

```{r}
race_summary <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean_english = mean(ENGLISH, na.rm = TRUE),
    median_english = median(ENGLISH, na.rm = TRUE),
    sd_english = sd(ENGLISH, na.rm = TRUE)
  )

race_summary
```

::: {.panel-tabset}

## The plot

```{r}
ggplot(race_summary,
       aes(x = reorder(RACE, median_english),
           y = median_english)) +
  geom_col(
    fill = "#2A6F6F",
    color = "#1F1F1F",
    linewidth = 0.35,
    width = 0.65
  ) +
  geom_text(
    aes(label = round(median_english, 1)),
    vjust = -0.4,
    size = 3.5
  ) +
  scale_y_continuous(
    name = "Median English score",
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Median English Score by Race",
    subtitle = "A simple summary chart to support the raincloud plot interpretation",
    x = "Race"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
race_summary <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean_english = mean(ENGLISH, na.rm = TRUE),
    median_english = median(ENGLISH, na.rm = TRUE),
    sd_english = sd(ENGLISH, na.rm = TRUE)
  )

ggplot(race_summary,
       aes(x = reorder(RACE, median_english),
           y = median_english)) +
  geom_col(
    fill = "#2A6F6F",
    color = "#1F1F1F",
    linewidth = 0.35,
    width = 0.65
  ) +
  geom_text(
    aes(label = round(median_english, 1)),
    vjust = -0.4,
    size = 3.5
  ) +
  scale_y_continuous(
    name = "Median English score",
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Median English Score by Race",
    subtitle = "A simple summary chart to support the raincloud plot interpretation",
    x = "Race"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

This extension is intentionally simple. The raincloud plot shows the full distribution, while the median bar chart gives a quick business-style summary. Used together, they make the interpretation clearer without making the exercise too complicated.

## Reference

- Introducing Ridgeline Plots
- Fundamentals of Data Visualization
- Raincloud plots: a multi-platform tool for robust data visualisation
- ggdist documentation