---
title: "Hands-on_Ex10"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
execute:
  warning: false
  message: false
---

# Information Dashboard Design: R methods

## Overview

A dashboard brings several compact views together so readers can scan status, compare performance and find exceptions quickly. In R, dashboard-style outputs can be built from data summaries, small plots, formatted tables and interactive tables.

In this hands-on exercise, the Coffee Chain dataset is used to create a small sales performance report. The supplied Microsoft Access database is kept in the data folder, but the exercise uses the prepared RDS version to avoid driver issues on macOS.

By the end of this exercise, we should be able to:

- import a prepared dashboard dataset;
- create performance summaries against budget;
- build bullet-chart style comparisons in `ggplot2`;
- draw small multiple sparklines;
- create a static report table with `gt`;
- create an interactive table with `reactable`.

## Installing and launching R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  lubridate,
  scales,
  gt,
  reactable,
  htmltools,
  knitr
)
```

## Importing Microsoft Access database

### The data set

The exercise folder contains two files:

- `Coffee Chain.mdb`, the original Microsoft Access database;
- `CoffeeChain.rds`, a prepared R object with the same working data used for this reproducible page.

Using the RDS file keeps the page portable because importing `.mdb` files on macOS can depend on external database drivers.

```{r}
coffee_path <- "data/rds/CoffeeChain.rds"

if (!file.exists(coffee_path)) {
  stop("CoffeeChain.rds is missing from data/rds.")
}

coffee <- readRDS(coffee_path) %>%
  mutate(
    Date = as.Date(Date),
    month = floor_date(Date, "month"),
    profit_gap = Profit - `Budget Profit`,
    sales_gap = Sales - `Budget Sales`,
    margin_gap = Margin - `Budget Margin`
  )

glimpse(coffee)
```

```{r}
coffee %>%
  head(8) %>%
  knitr::kable(format = "html")
```

### Data Preparation

The dashboard summaries compare actual sales and profit against budget by market.

```{r}
market_summary <- coffee %>%
  group_by(Market) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    `Budget Sales` = sum(`Budget Sales`, na.rm = TRUE),
    Profit = sum(Profit, na.rm = TRUE),
    `Budget Profit` = sum(`Budget Profit`, na.rm = TRUE),
    Margin = sum(Margin, na.rm = TRUE),
    `Budget Margin` = sum(`Budget Margin`, na.rm = TRUE),
    sales_gap = Sales - `Budget Sales`,
    profit_gap = Profit - `Budget Profit`,
    margin_gap = Margin - `Budget Margin`,
    .groups = "drop"
  ) %>%
  arrange(desc(Sales))

market_summary %>%
  knitr::kable(format = "html")
```

### Bullet chart in ggplot2

A bullet chart compares actual performance against a target in a compact form. Here, the bar shows actual sales and the vertical marker shows budget sales.

```{r fig.width=8, fig.height=5}
ggplot(market_summary, aes(y = fct_reorder(Market, Sales))) +
  geom_col(aes(x = Sales), fill = "#2A6F6F", width = 0.62) +
  geom_point(aes(x = `Budget Sales`), shape = 124, size = 9, colour = "#D95F02") +
  scale_x_continuous(labels = dollar) +
  labs(
    title = "Sales Against Budget by Market",
    subtitle = "Bars show actual sales; orange markers show budget sales",
    x = "Sales",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Plotting sparklines using ggplot2

### Preparing the data

```{r}
monthly_market <- coffee %>%
  group_by(Market, month) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    `Budget Sales` = sum(`Budget Sales`, na.rm = TRUE),
    Profit = sum(Profit, na.rm = TRUE),
    .groups = "drop"
  )
```

### Sparklines in ggplot2

Small multiple line charts act like dashboard sparklines. They preserve the trend for each market while using very little visual space.

```{r fig.width=10, fig.height=6}
ggplot(monthly_market, aes(month, Sales)) +
  geom_line(colour = "#2A6F6F", linewidth = 0.7) +
  geom_line(aes(y = `Budget Sales`), colour = "#D95F02", linetype = "dashed", linewidth = 0.55) +
  facet_wrap(~Market, scales = "free_y") +
  scale_y_continuous(labels = dollar_format(scale_cut = cut_short_scale())) +
  labs(
    title = "Monthly Sales Sparklines by Market",
    subtitle = "Solid line is actual sales; dashed line is budget sales",
    x = NULL,
    y = "Sales"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

## Static Information Dashboard Design: gt methods

The `gt` package is used to create a compact market performance table. Colour is applied to the gap columns so positive and negative performance can be scanned quickly.

```{r}
gt_market <- market_summary %>%
  gt() %>%
  tab_header(
    title = "Coffee Chain Market Performance",
    subtitle = "Actual performance compared with budget"
  ) %>%
  fmt_currency(
    columns = c(Sales, `Budget Sales`, Profit, `Budget Profit`, Margin, `Budget Margin`, sales_gap, profit_gap, margin_gap),
    currency = "USD",
    decimals = 0
  ) %>%
  data_color(
    columns = c(sales_gap, profit_gap, margin_gap),
    palette = c("#D95F02", "white", "#2A6F6F"),
    domain = NULL
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )

gt_market
```

## Interactive Information Dashboard Design: reactable methods

Interactive tables are useful when readers need sorting, filtering and compact comparison. The table below keeps the market-level view but allows sorting by any metric.

```{r}
sales_palette <- col_numeric(
  palette = c("#F2F0E6", "#2A6F6F"),
  domain = market_summary$Sales
)

gap_style <- function(value) {
  if (is.na(value)) {
    return(list())
  }

  if (value >= 0) {
    list(color = "#2A6F6F", fontWeight = "600")
  } else {
    list(color = "#D95F02", fontWeight = "600")
  }
}
```

```{r}
reactable(
  market_summary,
  searchable = TRUE,
  defaultPageSize = 6,
  columns = list(
    Sales = colDef(
      format = colFormat(currency = "USD", separators = TRUE, digits = 0),
      style = function(value) {
        list(background = sales_palette(value))
      }
    ),
    `Budget Sales` = colDef(format = colFormat(currency = "USD", separators = TRUE, digits = 0)),
    Profit = colDef(format = colFormat(currency = "USD", separators = TRUE, digits = 0)),
    `Budget Profit` = colDef(format = colFormat(currency = "USD", separators = TRUE, digits = 0)),
    Margin = colDef(format = colFormat(currency = "USD", separators = TRUE, digits = 0)),
    `Budget Margin` = colDef(format = colFormat(currency = "USD", separators = TRUE, digits = 0)),
    sales_gap = colDef(
      name = "Sales Gap",
      format = colFormat(currency = "USD", separators = TRUE, digits = 0),
      style = gap_style
    ),
    profit_gap = colDef(
      name = "Profit Gap",
      format = colFormat(currency = "USD", separators = TRUE, digits = 0),
      style = gap_style
    ),
    margin_gap = colDef(
      name = "Margin Gap",
      format = colFormat(currency = "USD", separators = TRUE, digits = 0),
      style = gap_style
    )
  )
)
```

## My Extension: Product performance view

The dashboard can also be viewed by product. This extension identifies the products with the largest profit gap.

```{r}
product_summary <- coffee %>%
  group_by(`Product Type`, Product) %>%
  summarise(
    Sales = sum(Sales, na.rm = TRUE),
    Profit = sum(Profit, na.rm = TRUE),
    `Budget Profit` = sum(`Budget Profit`, na.rm = TRUE),
    profit_gap = Profit - `Budget Profit`,
    .groups = "drop"
  ) %>%
  arrange(desc(profit_gap))
```

```{r fig.width=9, fig.height=6}
product_summary %>%
  slice_max(abs(profit_gap), n = 12, with_ties = FALSE) %>%
  mutate(Product = fct_reorder(Product, profit_gap)) %>%
  ggplot(aes(profit_gap, Product, fill = profit_gap >= 0)) +
  geom_col(width = 0.68) +
  scale_x_continuous(labels = dollar) +
  scale_fill_manual(values = c("TRUE" = "#2A6F6F", "FALSE" = "#D95F02"), guide = "none") +
  labs(
    title = "Largest Product Profit Gaps",
    x = "Profit minus budget profit",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Key Takeaways

- Dashboard views should combine summary, comparison and exception-finding.
- Bullet charts are compact ways to compare actual values against targets.
- Sparklines are useful when trends need to fit into a small space.
- Static `gt` tables are good for presentation, while `reactable` tables support exploration.
- A prepared RDS file is a practical reproducible alternative when Access drivers are unavailable.

## References

- R4VA chapter: <https://r4va.netlify.app/chap31>
