---
title: "Hands-on Exercise 3: Interactive Data Visualisation with R"
author: "Cheng Yuanyuan"
date: "2026-05-01"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

Interactive data visualisation is useful when a chart needs to do more than just “look nice”. A static chart can show the main pattern, but an interactive chart lets the reader slow down, hover, click, filter, and inspect the details behind the pattern.

In this exercise, I will use R to turn exam score data into interactive visualisations. The dataset includes students’ results for English, Mathematics and Science, so it is suitable for exploring score distributions, subject differences, and group patterns.

My focus is not to make the charts flashy. A chart that spins, blinks, and screams for attention is usually not smarter; it is just noisier. Instead, the interaction should have a clear job: revealing student-level details, supporting comparison, and making the chart easier to explore without overcrowding the page.

The guiding question for this exercise is:

> Can interaction help the reader understand the data more clearly, or is it just decoration with better marketing?

# Getting Started

Before building the interactive charts, I first load the R packages needed for this exercise. The main package is ggiraph, which allows ggplot charts to become interactive. I will also use plotly for browser-based interactive charts, DT for interactive tables, and tidyverse for data wrangling and static ggplot foundations.

In plain English, this step is like preparing the toolbox before touching the data. If the packages are not loaded properly, the rest of the exercise will complain dramatically.

```{r}
pacman::p_load(ggiraph, plotly, patchwork, DT, tidyverse)
```

# Importing Data

The dataset used in this exercise contains students’ examination results across three subjects: English, Mathematics and Science. Before creating any interactive visualisation, I first import the data and check whether the columns are read correctly.

This step looks boring, but it matters. If the data is imported wrongly, the interactive chart will only become an interactive mistake.

```{r}
  exam_data <- read_csv("../Hands-on_Ex02/Exam_data.csv")
```

```{r}
glimpse(exam_data)
```

# Data Preparation

Before drawing the interactive charts, I create a few helper columns. These columns are not changing the original meaning of the data. They simply make the later visualisations easier to read and easier to interact with.

The most important helper column is the tooltip text. Instead of showing only one value when the mouse hovers over a point, the tooltip will show the student ID, class, gender, race, and subject scores. This makes the chart less blind: the reader can inspect individual students without overcrowding the whole plot.

```{r}
exam_data <- exam_data %>%
  mutate(
    tooltip = paste0(
      "Student: ", ID,
      "\nClass: ", CLASS,
      "\nGender: ", GENDER,
      "\nRace: ", RACE,
      "\nEnglish: ", ENGLISH,
      "\nMaths: ", MATHS,
      "\nScience: ", SCIENCE
    ),
    total_score = ENGLISH + MATHS + SCIENCE,
    average_score = round(total_score / 3, 1)
  )
```

```{r}
exam_data %>%
  select(ID, CLASS, GENDER, RACE, ENGLISH, MATHS, SCIENCE, total_score, average_score, tooltip) %>%
  head()
```

# Interactive Maths Score Distribution

The first interactive chart focuses on Mathematics scores. A normal histogram or dot plot can show the overall score distribution, but it hides the individual students behind each dot. This is where interaction becomes useful.

In this chart, each dot represents one student. When the reader hovers over a dot, the tooltip reveals the student's class, gender, race, and scores across the three subjects. This keeps the chart clean while still allowing detailed inspection when needed.

The purpose is not to turn the chart into a toy. The purpose is to let the reader ask: who exactly is behind this point?

```{r}
p1 <- ggplot(
  data = exam_data,
  aes(x = MATHS)
) +
  geom_dotplot_interactive(
    aes(
      tooltip = tooltip,
      data_id = ID
    ),
    binwidth = 2.5,
    stackgroups = TRUE,
    method = "histodot",
    dotsize = 0.6,
    fill = "#2A6F6F",
    colour = "#1F1F1F",
    linewidth = 0.25
  ) +
  labs(
    title = "Distribution of Mathematics Scores",
    subtitle = "Hover over each dot to inspect student-level details",
    x = "Mathematics Score",
    y = "Number of Students"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

girafe(
  ggobj = p1,
  width_svg = 8,
  height_svg = 4
)
```

# Interactive English Score Distribution

The second chart looks at English scores. Instead of only showing the overall distribution, I separate the dots by gender. This makes the chart more useful for comparison because the reader can quickly check whether the score pattern looks balanced or whether one group appears more concentrated at certain score ranges.

The interaction remains simple: hover over a dot to inspect the student behind the score. This is a small design choice, but it prevents the chart from becoming a wall of labels.

```{r}
p2 <- ggplot(
  data = exam_data,
  aes(x = ENGLISH, fill = GENDER)
) +
  geom_dotplot_interactive(
    aes(
      tooltip = tooltip,
      data_id = ID
    ),
    binwidth = 2.5,
    stackgroups = TRUE,
    method = "histodot",
    dotsize = 0.6,
    colour = "#1F1F1F",
    linewidth = 0.25
  ) +
  scale_fill_manual(
    values = c(
      "Male" = "#2A6F6F",
      "Female" = "#E76F51"
    )
  ) +
  labs(
    title = "Distribution of English Scores by Gender",
    subtitle = "Hover over each dot to view student-level information",
    x = "English Score",
    y = "Number of Students",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

girafe(
  ggobj = p2,
  width_svg = 8,
  height_svg = 4
)
```

# Interactive Science Score Distribution

The third chart examines Science scores by class. Compared with the previous charts, this one shifts the focus from individual subject distribution to class-level comparison.

This is useful because classes may have different score patterns even when the overall subject distribution looks normal. By using class as the grouping variable, the chart gives a quick way to check whether some classes are clustered around higher or lower Science scores.

Again, the tooltip keeps the chart readable. The details are available when needed, but they do not attack the reader’s eyes all at once.

```{r}
p3 <- ggplot(
  data = exam_data,
  aes(x = SCIENCE, fill = CLASS)
) +
  geom_dotplot_interactive(
    aes(
      tooltip = tooltip,
      data_id = ID
    ),
    binwidth = 2.5,
    stackgroups = TRUE,
    method = "histodot",
    dotsize = 0.55,
    colour = "#1F1F1F",
    linewidth = 0.2
  ) +
  labs(
    title = "Distribution of Science Scores by Class",
    subtitle = "Hover over each dot to inspect individual student records",
    x = "Science Score",
    y = "Number of Students",
    fill = "Class"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

girafe(
  ggobj = p3,
  width_svg = 8,
  height_svg = 4.5
)
```

# 

# Interactive Relationship between Maths and Science

After looking at each subject separately, I now compare Mathematics and Science scores in the same chart. This is useful because the two subjects may move together: students who perform well in Mathematics may also perform well in Science.

Each point represents one student. The x-axis shows the Mathematics score, while the y-axis shows the Science score. I use class as the colour grouping so that the chart can show both the subject relationship and possible class-level patterns.

The interaction is important here because scatterplots can become crowded very quickly. Instead of labeling every point and creating a visual traffic jam, the tooltip allows the reader to inspect one student at a time.

```{r}
p4 <- ggplot(
  data = exam_data,
  aes(x = MATHS, y = SCIENCE, colour = CLASS)
) +
  geom_point_interactive(
    aes(
      tooltip = tooltip,
      data_id = ID
    ),
    size = 3,
    alpha = 0.75
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 0.7,
    colour = "#1F1F1F"
  ) +
  labs(
    title = "Relationship between Mathematics and Science Scores",
    subtitle = "Hover over each point to inspect student-level details",
    x = "Mathematics Score",
    y = "Science Score",
    colour = "Class"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

girafe(
  ggobj = p4,
  width_svg = 8,
  height_svg = 5
)
```

# 

# Plotly Interactive Chart

The previous charts use ggiraph to add interaction to ggplot graphics. In this section, I use plotly, which provides a more web-style interactive experience. The chart can be zoomed, moved, and explored directly in the browser.

Here, I compare students’ average scores with their total scores. This may sound repetitive at first, because average score and total score are mathematically linked. However, the chart is still useful as a clean interactive overview: students with higher total scores naturally appear higher, and the colour grouping by class helps reveal whether certain classes contain more high-performing or low-performing students.

This is also a good reminder that interactivity does not fix weak chart logic. A chart still needs a clear question. The question here is: which students and classes are sitting at the higher or lower end of overall performance?

```{r}
plot_ly(
  data = exam_data,
  x = ~average_score,
  y = ~total_score,
  color = ~CLASS,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  marker = list(
    size = 10,
    opacity = 0.75,
    line = list(
      width = 1,
      color = "#1F1F1F"
    )
  )
) %>%
  layout(
    title = "Overall Student Performance by Class",
    xaxis = list(title = "Average Score"),
    yaxis = list(title = "Total Score"),
    legend = list(title = list(text = "Class"))
  )
```

plot_ly( data = exam_data, x = \~average_score, y = \~total_score, color = \~CLASS, type = "scatter", mode = "markers", text = \~tooltip, hoverinfo = "text", marker = list( size = 10, opacity = 0.75, line = list( width = 1, color = "#1F1F1F" ) ) ) %\>% layout( title = "Overall Student Performance by Class", xaxis = list(title = "Average Score"), yaxis = list(title = "Total Score"), legend = list(title = list(text = "Class")) )

# Plotly Bubble Chart: Comparing Three Subjects

A scatterplot can compare two variables well, but the dataset has three subject scores. To avoid making three separate charts again, I use a bubble chart to place three subjects into one view.

In this chart, English score is shown on the x-axis, Mathematics score is shown on the y-axis, and the bubble size represents Science score. This makes the chart more meaningful than comparing total score with average score, because each visual element now carries a different piece of information.

The chart is not perfect. Bubble size is harder to compare precisely than position, so it should be used for general pattern detection rather than exact reading. Still, it works well for spotting students who perform consistently well across all three subjects, or students whose performance is uneven across subjects.

```{r}
plot_ly(
  data = exam_data,
  x = ~ENGLISH,
  y = ~MATHS,
  size = ~SCIENCE,
  color = ~CLASS,
  type = "scatter",
  mode = "markers",
  text = ~tooltip,
  hoverinfo = "text",
  sizes = c(8, 35),
  marker = list(
    opacity = 0.65,
    line = list(
      width = 1,
      color = "#1F1F1F"
    )
  )
) %>%
  layout(
    title = "Comparing English, Mathematics and Science Scores",
    xaxis = list(title = "English Score"),
    yaxis = list(title = "Mathematics Score"),
    legend = list(title = list(text = "Class"))
  )
```

# Interactive Data Table

Charts are useful for seeing patterns, but they are not always the best place to inspect exact values. For exact student records, an interactive table is cleaner and more practical.

In this section, I use DT to create a searchable and sortable table. This allows the reader to filter students by class, gender, race, or score values without forcing the chart to carry too much detail.

This is the less glamorous part of interactive visualisation, but it is often the part people actually use. A good table is not boring; it is just quietly doing the work.

```{r}
exam_data %>%
  select(
    ID, CLASS, GENDER, RACE,
    ENGLISH, MATHS, SCIENCE,
    total_score, average_score
  ) %>%
  datatable(
    filter = "top",
    options = list(
      pageLength = 10,
      autoWidth = TRUE
    ),
    rownames = FALSE,
    caption = "Interactive table of student examination results"
  )
```

# Key Observations

The interactive charts make the score patterns easier to inspect without overloading the page with labels. The dot plots show how the scores are distributed across the three subjects, while the scatterplots make it easier to compare relationships between subjects.

For Mathematics and Science, the scatterplot suggests that students who perform well in Mathematics often also perform well in Science. This is not surprising, because both subjects may involve similar reasoning and problem-solving skills. However, the interactive tooltip makes the chart more useful because individual student records can be checked directly instead of guessing from the point position.

The bubble chart adds another layer by showing English, Mathematics and Science scores in one view. This helps identify students with balanced performance across all three subjects. At the same time, bubble size should not be overread because size is less precise than x- and y-position.

The interactive table supports the charts by allowing exact records to be searched, filtered and sorted. This is useful when the reader wants to move from a general pattern to a specific student record.

# Conclusion

This exercise shows that interactive visualisation is most useful when it helps the reader explore details without making the chart messy. The hover tooltips are useful because they reveal student-level information only when needed. The plotly charts are useful because they allow zooming and direct exploration in the browser. The DT table is useful because it supports searching, sorting and filtering exact records.

The main lesson is that interaction should have a purpose. If interaction helps the reader inspect, compare or filter the data, it improves the visualisation. If it only adds movement or decoration, it becomes noise with better technology.

For this dataset, the interactive approach works well because the exam records contain both overall patterns and individual student details. A static chart can show the broad distribution, but an interactive chart makes it easier to move between the big picture and the specific student record.
