---
title: "Hands-on_Ex05c"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
---

# Visualising Uncertainty

## Learning Outcome

In this hands-on exercise, we focus on how uncertainty can be shown visually. A chart often gives a single value, such as a mean score, but that value is only an estimate. To make the analysis more complete, we also need to show how uncertain the estimate is.

This exercise uses error bars, confidence intervals, interactive plots, `ggdist` interval plots and hypothetical outcome plots. The main example uses Maths scores from the examination dataset.

By the end of this exercise, we should be able to:

- calculate summary statistics for point estimates;
- visualise standard error bars using `ggplot2`;
- visualise confidence intervals using `ggplot2`;
- create interactive uncertainty plots using `plotly`, `DT` and `crosstalk`;
- use `ggdist` to create interval-based uncertainty graphics;
- understand the basic idea of hypothetical outcome plots.

## Getting Started

### Installing and loading the packages

The packages used in this exercise support data wrangling, static plotting, interactive plotting, table display and uncertainty visualisation.

```{r}
pacman::p_load(plotly, crosstalk, DT,
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

### Data import

For this exercise, `Exam_data.csv` is used. The dataset contains student examination results and grouping variables such as race, gender and class.

```{r}
exam <- read_csv("Exam_data.csv")
```

```{r}
glimpse(exam)
```

```{r}
summary(exam$MATHS)
```

## Visualizing the uncertainty of point estimates: ggplot2 methods

A point estimate is a single summary value. For example, the mean Maths score of each race group is a point estimate. However, a point estimate alone can be misleading because it does not show how stable or uncertain that estimate is.

Uncertainty can be shown using standard error, confidence interval or other interval-based summaries.

::: {.callout-important}
Do not confuse uncertainty of a point estimate with the spread of the raw data. Standard deviation describes how spread out the observations are, while standard error describes how uncertain the estimated mean is.
:::

First, we calculate the number of students, mean, standard deviation and standard error of Maths scores by race.

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean = mean(MATHS, na.rm = TRUE),
    sd = sd(MATHS, na.rm = TRUE)
  ) %>%
  mutate(
    se = sd / sqrt(n - 1)
  )
```

::: {.panel-tabset}

## The table

```{r}
knitr::kable(my_sum, format = "html", digits = 2)
```

## The code chunk

```{r eval=FALSE}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean = mean(MATHS, na.rm = TRUE),
    sd = sd(MATHS, na.rm = TRUE)
  ) %>%
  mutate(
    se = sd / sqrt(n - 1)
  )

knitr::kable(my_sum, format = "html", digits = 2)
```

:::

The table gives the statistical foundation for the plots that follow. The value `se` will be used to draw uncertainty intervals around each group mean.

### Plotting standard error bars of point estimates

The first method is to draw standard error bars around the mean Maths score for each race group. The interval is calculated as:

`mean ± se`

This gives a compact view of how much uncertainty surrounds each mean estimate.

::: {.panel-tabset}

## The plot

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = RACE,
        ymin = mean - se,
        ymax = mean + se),
    width = 0.18,
    colour = "#1F1F1F",
    alpha = 0.9,
    linewidth = 0.55
  ) +
  geom_point(
    aes(x = RACE,
        y = mean),
    stat = "identity",
    color = "#2A6F6F",
    size = 2.2,
    alpha = 1
  ) +
  labs(
    title = "Standard Error of Mean Maths Score by Race",
    subtitle = "Error bars show mean ± standard error",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = RACE,
        ymin = mean - se,
        ymax = mean + se),
    width = 0.18,
    colour = "#1F1F1F",
    alpha = 0.9,
    linewidth = 0.55
  ) +
  geom_point(
    aes(x = RACE,
        y = mean),
    stat = "identity",
    color = "#2A6F6F",
    size = 2.2,
    alpha = 1
  ) +
  labs(
    title = "Standard Error of Mean Maths Score by Race",
    subtitle = "Error bars show mean ± standard error",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

The point shows the mean Maths score for each race group. The vertical line shows the uncertainty around that estimate. A longer error bar means the estimated mean is less precise.

### Plotting confidence interval of point estimates

Instead of using only standard error, we can show a 95% confidence interval. A common approximation is:

`mean ± 1.96 × se`

This interval is wider than the standard error bar because it is intended to show a plausible range for the population mean.

::: {.panel-tabset}

## The plot

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = reorder(RACE, -mean),
        ymin = mean - 1.96 * se,
        ymax = mean + 1.96 * se),
    width = 0.18,
    colour = "#1F1F1F",
    alpha = 0.9,
    linewidth = 0.55
  ) +
  geom_point(
    aes(x = reorder(RACE, -mean),
        y = mean),
    stat = "identity",
    color = "#D95F02",
    size = 2.2,
    alpha = 1
  ) +
  labs(
    title = "95% Confidence Interval of Mean Maths Score by Race",
    subtitle = "Groups are arranged by average Maths score",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = reorder(RACE, -mean),
        ymin = mean - 1.96 * se,
        ymax = mean + 1.96 * se),
    width = 0.18,
    colour = "#1F1F1F",
    alpha = 0.9,
    linewidth = 0.55
  ) +
  geom_point(
    aes(x = reorder(RACE, -mean),
        y = mean),
    stat = "identity",
    color = "#D95F02",
    size = 2.2,
    alpha = 1
  ) +
  labs(
    title = "95% Confidence Interval of Mean Maths Score by Race",
    subtitle = "Groups are arranged by average Maths score",
    x = "Race",
    y = "Mean Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

The confidence interval makes the uncertainty more visible. If one group has a small sample size, its interval may become much wider even if the mean looks high or low.

### Visualizing the uncertainty of point estimates with interactive error bars

Static plots are useful, but interactive plots allow users to inspect the values more directly. In this section, we combine `ggplot2`, `plotly`, `DT` and `crosstalk`.

A 99% confidence interval is used here:

`mean ± 2.58 × se`

::: {.panel-tabset}

## The plot

```{r}
shared_df <- SharedData$new(my_sum)

bscols(
  widths = c(4, 8),
  DT::datatable(
    shared_df,
    rownames = FALSE,
    class = "compact",
    width = "100%",
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    colnames = c("Race", "No. of pupils", "Avg. Scores", "Std Dev", "Std Error")
  ) %>%
    formatRound(
      columns = c("mean", "sd", "se"),
      digits = 2
    ),
  ggplotly(
    ggplot(shared_df) +
      geom_errorbar(
        aes(
          x = reorder(RACE, -mean),
          ymin = mean - 2.58 * se,
          ymax = mean + 2.58 * se
        ),
        width = 0.18,
        colour = "#1F1F1F",
        alpha = 0.9,
        linewidth = 0.55
      ) +
      geom_point(
        aes(
          x = RACE,
          y = mean,
          text = paste(
            "Race:", RACE,
            "<br>N:", n,
            "<br>Average Score:", round(mean, 2),
            "<br>99% CI: [",
            round(mean - 2.58 * se, 2), ", ",
            round(mean + 2.58 * se, 2), "]"
          )
        ),
        stat = "identity",
        color = "#2A6F6F",
        size = 2.3,
        alpha = 1
      ) +
      labs(
        title = "99% Confidence Interval of Mean Maths Score by Race",
        x = "Race",
        y = "Average Maths score"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
        panel.grid.minor = element_blank()
      ),
    tooltip = "text"
  )
)
```

## The code chunk

```{r eval=FALSE}
shared_df <- SharedData$new(my_sum)

bscols(
  widths = c(4, 8),
  DT::datatable(
    shared_df,
    rownames = FALSE,
    class = "compact",
    width = "100%",
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    colnames = c("Race", "No. of pupils", "Avg. Scores", "Std Dev", "Std Error")
  ) %>%
    formatRound(
      columns = c("mean", "sd", "se"),
      digits = 2
    ),
  ggplotly(
    ggplot(shared_df) +
      geom_errorbar(
        aes(
          x = reorder(RACE, -mean),
          ymin = mean - 2.58 * se,
          ymax = mean + 2.58 * se
        ),
        width = 0.18,
        colour = "#1F1F1F",
        alpha = 0.9,
        linewidth = 0.55
      ) +
      geom_point(
        aes(
          x = RACE,
          y = mean,
          text = paste(
            "Race:", RACE,
            "<br>N:", n,
            "<br>Average Score:", round(mean, 2),
            "<br>99% CI: [",
            round(mean - 2.58 * se, 2), ", ",
            round(mean + 2.58 * se, 2), "]"
          )
        ),
        stat = "identity",
        color = "#2A6F6F",
        size = 2.3,
        alpha = 1
      ) +
      labs(
        title = "99% Confidence Interval of Mean Maths Score by Race",
        x = "Race",
        y = "Average Maths score"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
        panel.grid.minor = element_blank()
      ),
    tooltip = "text"
  )
)
```

:::

This interactive version is useful because the viewer can inspect the mean, sample size and confidence interval directly. The table also gives the exact values behind the visualisation.

## Visualising Uncertainty: ggdist package

The `ggdist` package provides geoms and stats designed for visualising distributions and uncertainty. Compared with basic error bars, `ggdist` can show richer interval information in a more compact visual format.

In this section, Maths score is compared across race groups using several `ggdist` methods.

### Visualizing the uncertainty of point estimates: ggdist methods

The function `stat_pointinterval()` creates a point and interval plot. By default, it summarises the distribution and displays interval estimates.

::: {.panel-tabset}

## The plot

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    color = "#2A6F6F",
    point_size = 2.3,
    interval_size = 0.9
  ) +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Mean point with interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    color = "#2A6F6F",
    point_size = 2.3,
    interval_size = 0.9
  ) +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Mean point with interval plot",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

This plot is cleaner than a manually constructed error bar chart because the interval summary is handled directly by `ggdist`.

### Visualizing the uncertainty of median point estimates

We can also change the summary point from mean to median. This is useful when the distribution may be skewed or when we want a more robust measure of central tendency.

::: {.panel-tabset}

## The plot

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    .width = 0.95,
    .point = median,
    .interval = qi,
    color = "#D95F02",
    point_size = 2.3,
    interval_size = 0.9
  ) +
  labs(
    title = "Visualising Confidence Intervals of Median Maths Score",
    subtitle = "Median point with 95% interval",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    .width = 0.95,
    .point = median,
    .interval = qi,
    color = "#D95F02",
    point_size = 2.3,
    interval_size = 0.9
  ) +
  labs(
    title = "Visualising Confidence Intervals of Median Maths Score",
    subtitle = "Median point with 95% interval",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

Using median instead of mean gives a slightly different interpretation. It focuses on the middle student in each group rather than the arithmetic average.

### Visualizing 95% and 99% confidence intervals

A useful extension is to show more than one interval width in the same chart. Here, the chart shows both 95% and 99% intervals.

::: {.panel-tabset}

## The plot

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    .width = c(0.95, 0.99),
    color = "#2A6F6F",
    point_size = 2.4,
    interval_size = 0.9,
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising 95% and 99% Intervals of Maths Scores",
    subtitle = "Wider intervals represent greater uncertainty coverage",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(
    .width = c(0.95, 0.99),
    color = "#2A6F6F",
    point_size = 2.4,
    interval_size = 0.9,
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising 95% and 99% Intervals of Maths Scores",
    subtitle = "Wider intervals represent greater uncertainty coverage",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

The 99% interval is wider than the 95% interval because it covers a broader range of plausible values. This is the trade-off between confidence level and precision.

### Visualizing uncertainty with gradient interval

The function `stat_gradientinterval()` uses gradient colour to show distribution and uncertainty together. This gives the viewer a stronger visual cue about where the values are more concentrated.

::: {.panel-tabset}

## The plot

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "#2A6F6F",
    color = "#1F1F1F",
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Gradient interval plot by race",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "#2A6F6F",
    color = "#1F1F1F",
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising Confidence Intervals of Mean Maths Score",
    subtitle = "Gradient interval plot by race",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

This plot is more visually expressive than a simple error bar. It helps communicate that uncertainty is not just a hard boundary, but a range with different levels of density.

## Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

Hypothetical Outcome Plots, or HOPs, show uncertainty through animation. Instead of showing one fixed interval, the plot displays multiple possible outcomes over time.

This approach is helpful because it makes uncertainty feel more dynamic and less abstract.

### Installing ungeviz package

The original method uses the `ungeviz` package. If it has not been installed before, it needs to be installed from GitHub.

```{r eval=FALSE}
devtools::install_github("wilkelab/ungeviz")
```

::: {.callout-note}
This installation line is set to `eval=FALSE`, so it will not run during rendering. This prevents the website from failing if the package has not been installed yet.
:::

### Launch the application in R

After installation, the package can be loaded using `library()`.

```{r eval=FALSE}
library(ungeviz)
```

### Visualising Uncertainty with Hypothetical Outcome Plots (HOPs)

The code below shows how a HOP can be created using jittered points and animated hypothetical lines.

```{r eval=FALSE}
library(ungeviz)

ggplot(
  data = exam,
  aes(x = factor(RACE),
      y = MATHS)
) +
  geom_point(
    position = position_jitter(
      height = 0.3,
      width = 0.05
    ),
    size = 0.4,
    color = "#2A6F6F",
    alpha = 1 / 2
  ) +
  geom_hpline(
    data = sampler(25, group = RACE),
    height = 0.6,
    color = "#D95F02"
  ) +
  theme_bw() +
  labs(
    title = "Hypothetical Outcome Plot of Maths Scores by Race",
    subtitle = "Animated samples show uncertainty across possible outcomes",
    x = "Race",
    y = "Maths score"
  ) +
  transition_states(.draw, 1, 3)
```

Because animation and GitHub-based package installation may be unstable during website rendering, the HOP code is shown as a reference code chunk rather than executed directly.

### My Extension: Static uncertainty check

As a small extension, we can create a safer static version that shows the raw Maths scores together with the group mean. This does not replace HOPs, but it gives a clear and render-friendly way to show both individual variation and point estimates.

::: {.panel-tabset}

## The plot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = MATHS)) +
  geom_jitter(
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.2,
    color = "#2A6F6F"
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3,
    color = "#D95F02"
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.18,
    color = "#1F1F1F",
    linewidth = 0.55
  ) +
  labs(
    title = "Raw Maths Scores with Mean Confidence Intervals",
    subtitle = "Jittered points show sample variation, while error bars show uncertainty around the mean",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

## The code chunk

```{r eval=FALSE}
ggplot(exam,
       aes(x = RACE,
           y = MATHS)) +
  geom_jitter(
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 1.2,
    color = "#2A6F6F"
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 3,
    color = "#D95F02"
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.18,
    color = "#1F1F1F",
    linewidth = 0.55
  ) +
  labs(
    title = "Raw Maths Scores with Mean Confidence Intervals",
    subtitle = "Jittered points show sample variation, while error bars show uncertainty around the mean",
    x = "Race",
    y = "Maths score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 9),
    panel.grid.minor = element_blank()
  )
```

:::

This extension is intentionally simple. It makes the difference between sample variation and uncertainty around the mean easier to see.

## Reference

- ggplot2 documentation
- ggdist documentation
- plotly documentation
- DT documentation
- crosstalk documentation
- ungeviz package
