---
title: "Hands-on_Ex09c"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
editor: visual
execute:
  warning: false
  message: false
---

# Analytical Mapping

## Overview

Analytical mapping goes beyond displaying a single count. It uses derived measures, classification rules and outlier logic to support comparison across areas.

In this hands-on exercise, Nigerian water point data is used to map non-functional water points. The exercise starts with a count map, then moves to rate maps, percentile maps and box maps.

By the end of this exercise, we should be able to:

- import a prepared `sf` object from an RDS file;
- derive functional and non-functional rates;
- map counts and rates;
- build percentile classes;
- create a box-map classification based on the interquartile range;
- identify areas that are unusually high or low.

## Installing and launching R packages

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = "https://cloud.r-project.org")
}

pacman::p_load(
  tidyverse,
  sf,
  scales,
  knitr
)
```

## Importing data

```{r}
nga_path <- "data/rds/NGA_wp (1).rds"

if (!file.exists(nga_path)) {
  stop("NGA_wp (1).rds is missing from data/rds.")
}

nga_wp <- readRDS(nga_path) %>%
  st_make_valid()

glimpse(nga_wp)
```

```{r}
nga_wp %>%
  st_drop_geometry() %>%
  head(8) %>%
  knitr::kable(format = "html")
```

## Basic Choropleth Mapping

### Visualising distribution of non-functional water point

```{r fig.width=9, fig.height=7}
ggplot(nga_wp) +
  geom_sf(aes(fill = wp_nonfunctional), colour = "grey80", linewidth = 0.05) +
  scale_fill_gradient(
    low = "#F2F0E6",
    high = "#2A6F6F",
    labels = comma
  ) +
  labs(
    title = "Non-functional Water Points in Nigeria",
    fill = "Count"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

The count map highlights areas with more non-functional water points. However, areas with many total water points can naturally have larger counts, so rate maps are needed for fairer comparison.

## Choropleth Map for Rates

### Deriving proportion of functional and non-functional water points

```{r}
nga_rate <- nga_wp %>%
  mutate(
    pct_functional = if_else(total_wp > 0, wp_functional / total_wp, NA_real_),
    pct_nonfunctional = if_else(total_wp > 0, wp_nonfunctional / total_wp, NA_real_),
    pct_unknown = if_else(total_wp > 0, wp_unknown / total_wp, NA_real_)
  )
```

```{r}
nga_rate %>%
  st_drop_geometry() %>%
  select(ADM2_EN, ADM1_EN, total_wp, wp_functional, wp_nonfunctional, pct_nonfunctional) %>%
  arrange(desc(pct_nonfunctional)) %>%
  head(10) %>%
  mutate(pct_nonfunctional = percent(pct_nonfunctional, accuracy = 0.1)) %>%
  knitr::kable(format = "html")
```

### Plotting map of rate

```{r fig.width=9, fig.height=7}
ggplot(nga_rate) +
  geom_sf(aes(fill = pct_nonfunctional), colour = "grey80", linewidth = 0.05) +
  scale_fill_gradient(
    low = "#F2F0E6",
    high = "#D95F02",
    labels = percent,
    na.value = "grey90"
  ) +
  labs(
    title = "Non-functional Water Point Rate",
    subtitle = "Rate is calculated as non-functional water points divided by total water points",
    fill = "Rate"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Extreme Value Maps

### Percentile Map

Percentile mapping highlights values at the tails of a distribution. The classes below separate the bottom 1 percent, bottom 10 percent, middle range, top 10 percent and top 1 percent.

```{r}
nga_percentile <- nga_rate %>%
  mutate(
    pct_rank_nonfunctional = percent_rank(pct_nonfunctional),
    percentile_class = case_when(
      is.na(pct_rank_nonfunctional) ~ NA_character_,
      pct_rank_nonfunctional <= 0.01 ~ "Bottom 1%",
      pct_rank_nonfunctional <= 0.10 ~ "Bottom 10%",
      pct_rank_nonfunctional < 0.90 ~ "Middle 80%",
      pct_rank_nonfunctional < 0.99 ~ "Top 10%",
      TRUE ~ "Top 1%"
    ),
    percentile_class = factor(
      percentile_class,
      levels = c("Bottom 1%", "Bottom 10%", "Middle 80%", "Top 10%", "Top 1%")
    )
  )
```

```{r fig.width=9, fig.height=7}
ggplot(nga_percentile) +
  geom_sf(aes(fill = percentile_class), colour = "grey80", linewidth = 0.05) +
  scale_fill_manual(
    values = c(
      "Bottom 1%" = "#2A6F6F",
      "Bottom 10%" = "#8AC1B8",
      "Middle 80%" = "grey88",
      "Top 10%" = "#E6B655",
      "Top 1%" = "#D95F02"
    ),
    na.value = "grey95"
  ) +
  labs(
    title = "Percentile Map of Non-functional Water Point Rate",
    fill = "Class"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

### Box map

A box map applies the logic of a boxplot to a map. It highlights areas beyond the interquartile range as possible spatial outliers.

```{r}
rate_values <- nga_rate$pct_nonfunctional
q1 <- quantile(rate_values, 0.25, na.rm = TRUE)
q2 <- quantile(rate_values, 0.50, na.rm = TRUE)
q3 <- quantile(rate_values, 0.75, na.rm = TRUE)
iqr_value <- q3 - q1
lower_fence <- q1 - 1.5 * iqr_value
upper_fence <- q3 + 1.5 * iqr_value

nga_box <- nga_rate %>%
  mutate(
    box_class = case_when(
      is.na(pct_nonfunctional) ~ NA_character_,
      pct_nonfunctional < lower_fence ~ "Low outlier",
      pct_nonfunctional < q1 ~ "Lower quartile",
      pct_nonfunctional < q2 ~ "Below median",
      pct_nonfunctional < q3 ~ "Above median",
      pct_nonfunctional <= upper_fence ~ "Upper quartile",
      TRUE ~ "High outlier"
    ),
    box_class = factor(
      box_class,
      levels = c(
        "Low outlier",
        "Lower quartile",
        "Below median",
        "Above median",
        "Upper quartile",
        "High outlier"
      )
    )
  )
```

```{r fig.width=9, fig.height=7}
ggplot(nga_box) +
  geom_sf(aes(fill = box_class), colour = "grey80", linewidth = 0.05) +
  scale_fill_manual(
    values = c(
      "Low outlier" = "#1B4F72",
      "Lower quartile" = "#76B7B2",
      "Below median" = "#DDEBE9",
      "Above median" = "#F5D782",
      "Upper quartile" = "#E58E26",
      "High outlier" = "#B03A2E"
    ),
    na.value = "grey95"
  ) +
  labs(
    title = "Box Map of Non-functional Water Point Rate",
    subtitle = "High outliers are above Q3 plus 1.5 times the IQR",
    fill = "Box-map class"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## My Extension: Rate versus denominator

Rate maps can exaggerate small denominators. The scatterplot below checks whether high non-functional rates are attached to areas with very few total water points.

```{r fig.width=8, fig.height=5}
nga_rate %>%
  st_drop_geometry() %>%
  filter(total_wp > 0) %>%
  ggplot(aes(total_wp, pct_nonfunctional)) +
  geom_point(alpha = 0.45, colour = "#2A6F6F") +
  geom_smooth(method = "loess", se = FALSE, colour = "#D95F02") +
  scale_x_log10(labels = comma) +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Checking Rate Stability",
    subtitle = "Small denominators can create very high or very low rates",
    x = "Total water points, log scale",
    y = "Non-functional rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Key Takeaways

- Count maps and rate maps answer different questions.
- Percentile maps are useful for highlighting distribution tails.
- Box maps apply outlier logic to spatial data.
- Denominators matter: a high rate in a very small area should be interpreted carefully.

## References

- R4VA chapter: <https://r4va.netlify.app/chap23>

